---
title: "Intro to R - Part 3"
description: |
  Data Structures
author:
  - name: Michal Michalski
    url: https://dispersal.rbind.io/about.html
    affiliation: Durham University
date: 01-06-2021
repository_url: https://github.com/topographos/dispersal/
creative_commons: CC BY-NC
bibliography: intro_to_r.bib
categories:
  - R
  - Training
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 2
draft: false
---

# Data Structures

### Introduction

Most common dataset format is spreadsheet or a CSV file. Let's start by reading in a file of archaeological surveys from the Middle East.
We will load and save the data as an object named `surveys` ^[This will become a hover-able footnote]:

```{r}

surveys = read.csv("data/surveys.csv")

```

Let's have a look at the table:

```{r echo=FALSE, layout="l-body"}
library(knitr)
kable(head(surveys))
```



Now, we can jump to exploring the dataset right away, pulling out columns using the `$` operator:

```{r, echo=TRUE}

surveys$country
surveys$size_km

```

We can try other operations on the columns, for example if we discovered that each survey was actually 20 square kilometers bigger:

```{r}

surveys$size_km + 20

```

But what about adding values of the below columns

```{r echo=TRUE, message=FALSE, error=TRUE}

surveys$size_km + surveys$country

```

Understanding what happened here is key to successfully analyzing data in R.

The previous command produced an error because we tried to add for example `500` to `Syria`, a two different data types.
Thus, to avoid this, we can  check the data type of something by typing:

```{r}
class(surveys$size_km)
```

In R everything is an object.
Objects are used to hold data values.
The type of stored value is defined by data types.

There are six basic types in R:

* character
* numeric (real and decimal)
* integer
* logical
* complex
* raw

These data types can be combined to create data structures:

* atomic vector
* factor
* list
* matrix
* array
* data frame

R provides many functions to examine features of the objects, for example

* class() - what kind of object is it (high-level)?
* typeof() - what is the objects data type (low-level)?
* length() - how long is it? What about two dimensional objects?
* attributes() - does it have any metadata?

## Data Types

### Character

What is the site name?

```{r}
class("Tell Brak")
```

Text or string are called characters.

### Numeric

What is the site area?

```{r}
class(40.1)
```

Decimal values are called numeric.

### Integer

What is the site height?

```{r}
class(40L) # The L suffix forces the number to be an integer
```

Integer, a number written without a fractional component.In R all integers are also numeric.

### Logical

Is it a mound? (YES/NO == TRUE/FALSE)

```{r}
is_mound = TRUE
```

Logical values are denoted by Boolean values, TRUE or FAlSE.

## Data Structures
### Vectors

Vector are one-dimensional arrays that hold one type of data.

```{r}
my_vector = vector(length = 3)
my_vector
```

A vector in R is essentially an ordered list of things, with the special condition that everything in the vector must be the same basic data type. If you do not choose the data type, it will default to `logical`; or, you can declare an empty vector of whatever type you like.


```{r}
another_vector = vector(mode='character', length = 3)

another_vector

str(another_vector)

```

The somewhat cryptic output from this command indicates the basic data type found in this vector - in this case `chr`, character; an indication of the number of things in the vector - actually, the indexes of the vector, in this case [1:3]; and a few examples of what’s actually in the vector - in this case empty character strings. If we similarly do

```{r}
str(surveys$sites_number)
```

We see that `surveys$sites_number` is a vector, too - the columns of data we load into R data frames are all vectors, and that’s the root of why R forces everything in a column to be the same basic data type.

You can also make vectors with explicit contents with the combine function:

```{r}
combine_vector = c(178, 229, 31,42)
combine_vector
```

### Factors
### Lists



## Functions Review
::: l-body-outset
| Function         | Description                                    |
|------------------|------------------------------------------------|
| `read.table`     | Maximum rows to print per page.                |
| `read.csv`       | Maximum rows in the table (defaults to 1000).  |
| `read.delim`     | Maximum columns in the table (defaults to 10). |
| `class()`        | Print row names as part of the table.          |
| `str()`          | returns structure of the object                |
| `vector()`       | produces a vector of defined mode and length   |
| `c()`            | combine function                               |
:::




